---
title: "Data 608 HW1"
author: "Amit Kapoor"
date: "9/3/2020"
output:
  pdf_document: default
  html_notebook: default
  html_document: default
---

**Principles of Data Visualization and Introduction to ggplot2**

```{r message = FALSE, warning = FALSE}
library(dplyr)
library(ggplot2)
```


I have provided you with data about the 5,000 fastest growing companies in the US, as compiled by Inc. magazine. lets read this in:

```{r}

inc <- read.csv("https://raw.githubusercontent.com/charleyferrari/CUNY_DATA_608/master/module1/Data/inc5000_data.csv", header= TRUE)

```

And lets preview this data:

```{r}

head(inc)

```
```{r}

summary(inc)

```

Think a bit on what these summaries mean. Use the space below to add some more relevant non-visual exploratory information you think helps you understand this data:

```{r glimpse, echo=TRUE}

# Insert your code here, create more chunks as necessary
glimpse(inc)
```


```{r ind, echo=TRUE}

# Count by industries
industry <- inc %>% 
  group_by(Industry) %>% 
  count(Industry) %>% 
  arrange(desc(n))

industry
```


```{r rev, echo=TRUE}

# select top 5 industries by revenue
ind_revenue <- inc %>% 
  group_by(Industry) %>% 
  summarise(tot_rev_ind = sum(Revenue)) %>% 
  mutate(total_revenue_billions = round((tot_rev_ind / 1e9), 1)) %>% 
  select(-tot_rev_ind) %>%
  arrange(desc(total_revenue_billions)) %>% 
  top_n(n = 5)

ind_revenue
```


```{r employs, echo=TRUE}

# select top 5 industries that employs most
ind_employs <- inc %>% 
  group_by(Industry) %>% 
  summarise(tot_ind_emp = sum(Employees)) %>% 
  arrange(desc(tot_ind_emp)) %>% 
  top_n(n = 5)

ind_employs
```



## Question 1

Create a graph that shows the distribution of companies in the dataset by State (ie how many are in each state). There are a lot of States, so consider which axis you should use. This visualization is ultimately going to be consumed on a 'portrait' oriented screen (ie taller than wide), which should further guide your layout choices.

```{r ques1, echo=TRUE, warning = FALSE}

# Answer Question 1 here

# the distribution of companies by State
inc %>% count(State) %>% 
  ggplot(aes(x=reorder(State, n), y=n, fill=n)) + 
  geom_col() + 
  coord_flip() + 
  xlab("States") +
  ylab("Number of Companies") +
  ggtitle("Number of Companies by state")
           

```



## Quesiton 2

Lets dig in on the state with the 3rd most companies in the data set. Imagine you work for the state and are interested in how many people are employed by companies in different industries. Create a plot that shows the average and/or median employment by industry for companies in this state (only use cases with full data, use R's `complete.cases()` function.) In addition to this, your graph should show how variable the ranges are, and you should deal with outliers.

```{r ques2, echo=TRUE, warning = FALSE}

# Answer Question 2 here

# consider complete cases
inc_complete <- inc[complete.cases(inc),]

# for NY state
inc_complete %>% 
  filter(State=="NY") %>% 
  ggplot(aes(x=Industry, y=Employees)) + 
  geom_boxplot(width=.5, fill="grey", outlier.colour=NA) +
  stat_summary(aes(colour = "mean"), fun.y = mean, geom="point", fill="black", colour="red", shape=21, size=2, show.legend=TRUE) +
  stat_summary(aes(colour = "median"), fun.y = median, geom="point", fill="blue", colour="blue", shape=21, size=2, show.legend=TRUE) +
  coord_flip(ylim = c(0, 1500), expand = TRUE) +   
  scale_y_continuous(labels = scales::comma, breaks = seq(0, 1500, by = 100)) +
  xlab("Industry") +
  ylab("Employees by industry for companies") +
  ggtitle("Mean and Median Employment by Industry in NY State") + 
  theme(panel.background = element_blank(), legend.position = "top")

```




## Question 3

Now imagine you work for an investor and want to see which industries generate the most revenue per employee. Create a chart that makes this information clear. Once again, the distribution per industry should be shown.

```{r ind-rev-per-emp, echo=TRUE, warning = FALSE}

# Answer Question 3 here

# group by industry and calculate revenue ind per employee
ind_rev_per_emp <- inc[complete.cases(inc),] %>%
  group_by(Industry) %>%
  summarise(Rev_ind_per_emp=sum(Revenue) / sum(Employees))  %>%
  arrange(desc(Rev_ind_per_emp))

ind_rev_per_emp
```



```{r ques3, echo=TRUE, warning = FALSE}

# plot industries that generate the most revenue per employee
ggplot(ind_rev_per_emp, aes(x=reorder(Industry, Rev_ind_per_emp), y=Rev_ind_per_emp)) + 
  geom_bar(stat = 'Identity') +
  coord_flip() +
  xlab("Industries") +
  ylab("Revenue per employee") +
  ggtitle("Industries revenue per employee") +
  scale_y_continuous(labels = scales::comma)
```
















